Crystal as a thermodynamic computer

### 2024-09-14 paper/main.tex
shorten section D, added main result

\jinguo{
    Need to introduce the heat zone.
}
\ym{
    I tried. Please check if it is clear and if i still need to cut some details off.
}

To overcome the overlap gap property, we propose to cool the system piece by piece, by introducing a heat zone with width $W$.
It represents the region of the system that interacts with the heat bath, as shown in \Cref{Surface-programmable-thermodynamic computer}.
Let layer $i$ be defined as the set of spins with x-coordinate $i$.
The system is devided into $\lceil\frac{M}{W}\rceil$ parts,
where part $j$ consists of layers in the range $[jW, (j+1)W-1]$.
The system can then be cooled by sequentially cooling each parts with an idealized heat bath, 
which maintains a space-invariant temperature within the heat zone while keeping other regions at zero temperature.
To drive the system toward the ground state, the cooling order should proceed from left to right
, corresponding to the computation direction of the cellular automata.

Assume the annealing procedure in each part has an error probability $\epsilon$ that prevents at least one layer
from reaching the local ground state. 
Repeating the procedure $\frac{M}{W}$ times results in a success probability of $(1 - \epsilon)^{\frac{M}{W}}$ for finding the global ground state. 
To keep this probability constant as $M \to \infty$, the error probability $\epsilon$ should scale as $\frac{W}{M}$.

Now we turn to determining the sweep time $t_{th}$ required to achieve an error probability $\epsilon$ in each part. 
Let $\lambda_1$ and $\lambda_2$ be the largest and second largest eigenvalues of the Markov transition matrix, respectively. 
We assume the relationship between $\frac{\lambda_2}{\lambda_1}$ and $W$ can be expressed as
\begin{equation}
    \label{eq:spectral-gap}
    1 - \frac{\lambda_2}{\lambda_1} \sim e^{-W}.
\end{equation}
Consequently, $t_{th}$ is related to $\epsilon$ by:
\begin{equation}
    \left( \frac{\lambda_2}{\lambda_1} \right) ^ {t_{th}} = \epsilon \rightarrow t_{th} \sim \ln(\frac{M}{W})e^{W},
\end{equation}
and the total sweep times scale as $\frac{M}{W}\ln(\frac{M}{W})e^W$.

@@

In practice, temperature changes cannot occur instantaneously, and the temperature distribution within the heating zone may not be uniform.
Thus, we come up with a more realistic model for the block annealing process.

Now suppose the heat bath only connects to a small rectangle region of the system, creating a temperature gradient through heat flux,
The temperature $T(x, t)$  at a position with x-coordinate  $x$  and time  $t$  can be modeled as  $T(x, t) \propto \Lambda^{-|x - L(t)|}$ , 
where $\Lambda$ is a heat flux constant and $L(t)$ denotes the center of the heat bath at time $t$.

The block annealing process then can be reconstruct as moving
the heat bath from the left surface to the right surface at speed $v$. Defining $\Delta E_{max}$ to be the
largest possible energy change when a spin is fliped.
The effective heat zone with width $W'$ should then be $[y-W', y]$, where position $y$ satisfying 
$\exp(\frac{\Delta E_{max}}{T(y)}) = 0.1$ and $y<L(t)$. Since the layers
in fronts of the effective heat zone can only be fliped with a $\epsilon$ probability if the local energy would increase. 
The effective width $W'$ then should scale as $\exp(-\Delta E_{max} \Lambda^{cW'}) \sim \epsilon \sim \frac{W'}{M}$, where $c$ is an empirical constant for effective width and can be determined during the experiment. 
\ym{I am not sure about how to give a concrete definition of the effective width $W'$.}

The total sweep time with a fixed inputs size $N$ can be calculated as

\begin{equation}
    \label{eq:total-annealing-time}
    t_{total} \propto \frac{M}{W'}(\ln(\frac{M}{W'}))^{1+ \frac{1}{c\ln(\Lambda)}}
\end{equation}

@@

\subsubsection{Testing model with non-equilibrium simulated annealing}\label{some-details-about-non-equilibrium-simulated-annealing}\jinguo{Suggested subtitle: Simulating temperature gradient}

When we flip one spin, there would be some energy difference
$\Delta E_{i}$ in field $i$ with temperature $T_i$. The
probability to accept this flip would be

@@

It is easily to observe that when the system returns to
heat-equilibrium, the probability reverts to the classical HeatBath
acceptance probability. We also provided a more general proof by constructing another equivalent model in \Cref{proof-of_non-equilibrium-SA}.

Given that simulating a CA with a deep computational layer on the Ising model requires many spins, making large-scale numerical simulations impractical, we opted to introduce a similar but simpler energy model to observe the behavior on a larger scale.
Let us define a simpler but ``equivalent'' energy model on a 2D lattice of size $N \times M$, whose hamiltonian
is given by
\begin{equation}\label{eq:toy-hamiltonian}
E(\vn) = \sum_{j=1}^{M-1}\sum_{i=1}^{N} (n_{i+1, j} \land \neg(n_{i, j} \land \neg n_{i-1, j})) \veebar n_{i,j+1},
\end{equation}
where $n_{i, j} \in \{0, 1\}$ is the state of the cell at row $i$ and column $j$, the lattice is periodic in the space direction, i.e. $n_{0, t} = n_{N, j}$.
%$S$ to be the length of the material along the ``Space'' axis, $D$
%to be the length of the material along the ``Time'' axis. 
%Its ground state with energy $0$ also encodes a $110$ CA, where the direction of computation is from the left to right.

\jinguo{show a comparison from right to left}
\ym{I will do this in the next version.}

Here we conducted numerical experiments to examine how the cooling time $T_{c}$ scales with (i)
the ``Space'' dimension $N$, (ii) the ``Time'' dimension $M$, (iii) the ``Width'' of the heat zone $W$, and (iv) the temperature gradient $\Lambda$.
We use the number of SA sweeps as the metric to measure the cooling time, which could be determined with the binary search.
We keep changing the heat bath moving speed $v$ (in unit of columns/sweep) until the system can reach a ground state configuration with $\sim 50\%$ probability.

\begin{figure}[]
    \centering
    \includegraphics[width=\columnwidth,keepaspectratio]{../notes/images/main_result.pdf}
    \caption{Numerical Result
    \jinguo{images should use the pdf format, the font size should be larger. Is it possible to verify that the Ising model is similar to the toy model? What is $T$ in the xlabel?}
    \ym{I changed the labels. Maybe i need to conduct another numerical experiment
    to verify the similarity between the Ising model and the toy model.}
    }
    \label{fig:main-numerical-result}
\end{figure}

\jinguo{you need to explain the results... what can we learn from the results?}
\jinguo{bad variable name: Sweep Time}
\Cref{fig:main-numerical-result}(a) shows the relationship between
$\ln(\ln(M))$ and $\ln(v)$ shoule be linear, which implies 
the time complexity $T_c$ along the ``Time'' direction should scale as $M\ln(M)^{f(\Lambda)}$. 
We then fit the slopes of each line and $\frac{1}{\ln(\Lambda)}$ and find the 
empirical constant $c$ in \Cref{eq:total-annealing-time} to be $3$. \Cref{fig:main-numerical-result}(b) shows 
the the relationship between $T_c$ and $M$ should be linear. Combining these two results, 
we can conclude that the time complexity should scales as $NM\log(M)^{1+\frac{1}{3\ln(\Lambda)}}$, 
which is consistent with our theoretical prediction in \Cref{eq:total-annealing-time}.

>>>>

To overcome the overlap gap property, we propose to do block annealing, by introducing a heat zone with width $W$.
It represents the region of the system that interacts with the heat bath, as shown in \Cref{Surface-programmable-thermodynamic computer}.
Let layer $i$ be defined as the set of spins with x-coordinate $i$.
The system is devided into $\lceil\frac{M}{W}\rceil$ blocks,
where block $j$ consists of layers in the range $[(j-1)W+1, jW]$.
The system can then be cooled by sequentially cooling the blocks from left to right with an heat bath, where the temperature of the heat bath is gradually decreased.
We treat it as an ideal bath as it maintains a space-invariant temperature within the heat zone while keeping other regions at zero temperature.

Assume the SA procedure in each block has an error probability $\epsilon$.
Repeating the procedure in $\frac{M}{W}$ blocks results in a success probability of $(1 - \epsilon)^{\frac{M}{W}}$ for finding the global ground state. 
To keep this probability constant as $M \to \infty$, the error probability $\epsilon$ should scale as $\frac{W}{M}$.

Now we turn to determining the sweep time $t_{th}$ required to achieve $\epsilon$ error in each procedure. 
Let $\lambda_1$ and $\lambda_2$ be the largest and second largest eigenvalues of the Markov transition matrix, respectively. 
We assume the relationship between $\frac{\lambda_2}{\lambda_1}$ and $W$ can be expressed as
\begin{equation}
    \label{eq:spectral-gap}
    1 - \frac{\lambda_2}{\lambda_1} \sim e^{-W}.
\end{equation}
Consequently, $t_{th}$ is related to $\epsilon$ by:
\begin{equation}
    \left( \frac{\lambda_2}{\lambda_1} \right) ^ {t_{th}} = \epsilon \rightarrow t_{th} \sim \ln(\frac{M}{W})e^{W},
\end{equation}
and the total sweep times scale as $\frac{M}{W}\ln(\frac{M}{W})e^W$.

@@

In practice, the heat bath and heat flux should establish a temperature gradient in the system.
% Suppose the heat bath only connects to a small rectangle region of the system.
The temperature $T(x, t)$  at a position with x-coordinate  $x$  and time  $t$  can be modeled as  $T(x, t) \propto \Lambda^{-|x - L(t)|}$ , 
where $\Lambda$ is a heat flux constant and $L(t)$ denotes the center of the heat bath at time $t$.

The block annealing process then can be reconstruct as moving
the heat bath from the left surface to the right surface at speed $v$. Defining $\Delta E_{max}$ to be the
largest possible energy change when a spin is fliped.
The effective heat zone with width $W'(\Lambda)$ should then be $[y-W'(\Lambda), y]$, where position $y$ satisfying 
$\exp(\frac{\Delta E_{max}}{T(y)}) = 0.1$ and $y<L(t)$. 
% Since the layers in fronts of the effective heat zone can only be fliped with a $\epsilon$ probability if the local energy would increase. 
The effective width $W'$ then should scale as $\exp(-\Delta E_{max} \Lambda^{cW'(\Lambda)}) \sim \epsilon$, where $c$ is an empirical constant and can be determined during the experiment. 
\ym{I am not sure about how to give a concrete definition of the effective width $W'$.}

The total sweep time with a fixed inputs size $N$ can then be calculated as

\begin{equation}
    \label{eq:total-annealing-time}
    t_{total} \propto \frac{M}{W'(\Lambda)}(\ln(\frac{M}{W'(\Lambda)}))^{1+ \frac{1}{c\ln(\Lambda)}}
\end{equation}

@@

\subsection{Model and Method}
Given that simulating a CA with a deep computational layer on the Ising model requires many spins, making large-scale numerical simulations impractical, we opted to introduce a similar but simpler energy model to observe the behavior on a larger scale.
Let us define a simpler but ``equivalent'' energy model on a 2D lattice of size $N \times M$, whose hamiltonian
is given by
\begin{equation}\label{eq:toy-hamiltonian}
E(\vn) = \sum_{j=1}^{M-1}\sum_{i=1}^{N} (n_{i+1, j} \land \neg(n_{i, j} \land \neg n_{i-1, j})) \veebar n_{i,j+1},
\end{equation}
where $n_{i, j} \in \{0, 1\}$ is the state of the cell at row $i$ and column $j$, the lattice is periodic in the space direction, i.e. $n_{0, t} = n_{N, j}$.
%$S$ to be the length of the material along the ``Space'' axis, $D$
%to be the length of the material along the ``Time'' axis. 
%Its ground state with energy $0$ also encodes a $110$ CA, where the direction of computation is from the left to right.

When one spin is flipped in the SA, there would be some energy difference
$\Delta E_{i}$ in field $i$ with temperature $T_i$. 
We choose the probability to accept this flip to be

@@

It is easily to observe that when the system returns to
heat-equilibrium, the probability reverts to the Glauber dynamics.
We also provided a more general proof by constructing another equivalent model in \Cref{proof-of_non-equilibrium-SA}.

\subsection{Numerical Result}
\jinguo{show a comparison from right to left}

Here we conducted numerical experiments to examine how the cooling time $T_{c}$ scales with (i)
the ``Space'' dimension $N$, (ii) the ``Time'' dimension $M$, (iii) the ``Width'' of the heat zone $W$, and (iv) the temperature gradient $\Lambda$.
\ym{I have not conducted experiments for (iii) and (iv).}
We use the number of SA sweeps as the metric to measure the cooling time, which could be determined with the binary search.
We keep changing the heat bath moving speed $v$ (in unit of columns/sweep) until the system can reach a ground state configuration with $\sim 50\%$ probability.

\begin{figure*}[]
    \centering
    \vspace*{-\topskip}
    \includegraphics[width=\textwidth,keepaspectratio]{../notes/images/main_result.pdf}
    \caption{Numerical Result.
    \jinguo{images should use the pdf format, the font size should be larger. Is it possible to verify that the Ising model is similar to the toy model? What is $T$ in the xlabel?}
    \ym{If a comparison between Ising model and toy model is a must, plese inform me.}
    }
    \label{fig:main-numerical-result}
\end{figure*}

% \jinguo{you need to explain the results... what can we learn from the results?}
\Cref{fig:main-numerical-result}(a) shows the relationship between
$\ln(\ln(M))$ and $\ln(v)$ shoule be linear, which implies 
the time complexity $T_c$ along the ``Time'' direction should scale as $M\ln(M)^{f(\Lambda)}$. 
We then fit the slopes of each line and $\frac{1}{\ln(\Lambda)}$ and find the 
empirical constant $c$ in \Cref{eq:total-annealing-time} to be $3$.

\Cref{fig:main-numerical-result}(b) shows 
the the relationship between $T_c$ and $N$ should be linear. Combining these two results, 
we can conclude that the time complexity should scales as $NM\log(M)^{1+\frac{1}{3\ln(\Lambda)}}$, 
which is consistent with our theoretical prediction in \Cref{eq:total-annealing-time}.

\Cref{fig:main-numerical-result}(c) compares the time complexity with different annealing
methods. It costs exponential time to cool the system with the global annealing method no matter
we set the system to perform a deterministic or non-deterministic computation. Using the blocking
annealing method to perform a non-deterministic computation is still hard, too.
